RVD2: An ultra-sensitive variant detection model for low-depth targeted next-generation sequencing data

### 2013-12-04 doc/bioinformatics2013/RVD2.tex
updated TODO in manuscript

\begin{document}
%TODO: Label all results/figures with comment indicating what experiment folder those results depend on and come from.

\begin{abstract}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples. We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over wide ranges of median read depth and a minor allele frequency. We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

@@

RVD uses a two-stage approach for detecting for rare variants. First, it estimates the parameters of two hierarchical Bayesian models; one using data from the sample of interest (case) and one using data from a known reference sample (control). Then, it tests for a significant difference between model parameters in the case and control samples and returns called variant positions.

For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$. The generative process is as follows:

@@

This process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision. The global error rate, $\mu_0$, estimates the expected error rate across all locations. The global precision, $M_0$, estimates the variation in the error rate  across locations. The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.

@@

Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model. In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\cite{}. A rounded box or ``plate" represents replication of the nodes within the plate. This graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.

@@

\begin{equation}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right)
\end{equation}

The individual conditional distributions are:
\begin{align}
p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
\end{align}

Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data at location $j$ in replicate $i$, 
\begin{equation}
p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{equation}

Finally, the complete log-likelihood for the data set is

@@

The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution. However, there is not a closed form for the product of a Beta distribution with another Beta distribution and exact inference is intractable.  exact inference is intractable.

@@

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. 
The estimates for the position-level parameters are 
$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ 
and 
$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. 
The estimates for the genome-level parameters are 
$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ 
and 
$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encounter such pathological behavior in this application.

@@

%%%%%%%%%%%%%%%%%%
% Sampling mu
%%%%%%%%%%%%%%%%%%
\subsection{Sampling from $p \left( \mu_j |\theta_{ji},\mu_0,M_0\right)$}
The posterior distribution over $\mu_j$ given its Markov blanket is 
\begin{equation}
	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
\end{equation}

@@

A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$. The acceptance probability is then
\begin{equation}
	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
\end{equation}

We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise. Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position.

We resample from the proposal if the sample is outside of the support of the posterior distribution. We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples. Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this algorithm can be distributed across $J$ processors.

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Posterior Density Test}\label{sec:hypothesis_test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ for both the control and case data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

@@

The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ maximizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
\begin{equation}
	\tau^* = \arg\max_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
\end{equation}

While we are able to compute the optimal $\tau$ threshold for this test data set, in general we would not have access to $\tau^*$. With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest. Absent this information we set $\tau = 0$.

@@

The non-reference bases may be due to a true mutation or due to a random sequencing error and we would like to differentiate these two scenarios. We assume non-reference read counts caused by a non-biological mechanism results in a uniform distributed among three non-reference categories. In contrast, the distribution of counts among three non-reference bases caused by biological mutation would be biased to one base.

@@

where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. Cressie and Read (1984) proved that when $\lambda = 2/3$ this statistic is XXX.

At each position, we have one statistic for each replicate. We use Fisher's combined probability test to combine the $N$ p-values into a single p-value for each position,

\begin{equation}\label{eqn:fisher_combined}
	X^2 = -2 \sum_{i=1}^N \ln(p_i).
\end{equation}

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2*N$ degrees of freedom when the null hypothesis is true.

Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) across positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test}.

@@

We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms. The synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions. The HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.

@@

Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\cite{}. We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood. 
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we only used the Bayesian test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC}B-C for larger MAFs show a similar relationship between coverage depth and accuracy.

@@

Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.

\begin{figure}[h]
\begin{center}
\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
\label{fig:MAF}
\end{center}
\end{figure}
% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in calling all of the true variants and none of the true negatives.

[DISCUSS \citet{Stead:2013fu}]

@@

The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample. The sensitivity increases considerably with read depth at a slight expense to specificity. With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs. However, in practice one may not know the optimal $\tau^*$ a-priori. With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives. When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases. The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see supplementary table SX).

@@

VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level. However, the FDR is high due to many false positives. Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and 1.00 for the highest read depth. Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF. However, it does not call any variants at or below 1\% MAF.  MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths. But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\%-100\% MAF.

@@

RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX. Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants. The two heterozygous variants had a loss-of-heterozygosity in the tumor sample.

Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are germline homozygous mutations. Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample. In the tumor cells we identified all five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T. Positions chr7:154754371T$>$C and chr7:154758813G$>$A then loss-of-heterozygosity events.

%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.
%TODO: place these rs labels beneath the position labels on the x-axis of the figure. Remove the above sentence and add. 
%Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding rs identities are shown in the figure.

@@

The original research describing this sample used Strelka to identify mutations in the same sample. They identified chr7:154760439 as variant, but did not call the other four variants. In particular strelka missed the two LOH events.

@@

We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data. Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths.

Our inference algorithm uses Gibbs sampling and Metropolis-Hastings sampling to estimate the empirical Bayes statistical model. This sampling procedure provides a guarantee to identify the global optimal parameter settings in the long run. However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches. We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.

@@

%%%%%%%%%%%%%%%%%%%
% Appendix B: Comparison Statistics
%%%%%%%%%%%%%%%%%%%
\section{Algorithm Comparison Statistics}\label{sec:app_comparison}

@@

%%%%%%%%%%%%%%%%%
% Appendix D: Parameter Settings for Variant Calling Algorithms
%%%%%%%%%%%%%%%%%
\section{Parameter settings for other variant calling algorithms}

>>>>

\begin{document}

\begin{abstract}
Next-generation sequencing technology is increasingly being used for clinical diagnostic tests. Unlike research cell lines, clinical samples are often genomically heterogeneous due to low sample purity or the presence of genetic subpopulations. However, many variant calling algorithms are optimized to call single nucleotide polymorphisms in homogeneous rather than heterogeneous samples. We present a novel variant calling algorithm that uses a hierarchical Bayesian model to estimate allele frequency and call variants in heterogeneous samples. We show that our algorithm improves upon current classifiers and has higher sensitivity and specificity over a wide range of median read depth and minor allele frequency. We identify five mutations in the PAXP1 gene in a matched clinical breast ductal carcinoma tumor sample; two of which are loss-of-heterozygosity events.

@@

Several studies have compared the relative performance of these algorithms.
%TODO: [DISCUSS \citet{Stead:2013fu}] and others

@@

RVD uses a two-stage approach for detecting for rare variants. First, it estimates the parameters of a hierarchical Bayesian model under two sequencing data sets: one from the sample of interest (case) and one from a known reference sample (control). Then, it tests for a significant difference between key model parameters in the case and control samples and returns called variant positions.

For a given sample, the observed data consists of two matrices $r \in \RR^{J \times N}$ and $n \in \RR^{J \times N}$, where $r_{ji}$ is the number of reads with a non-reference base at location $j$ in experimental replicate $i$ and $n_{ji}$ is the total number of reads at location $j$ in replicate $i$. The model generative process is as follows:

@@

The generative process involves several hyperparameters: $\mu_0$, a global error rate; $M_0$, a global precision; $M_j$, a local precision. The global error rate, $\mu_0$, estimates the expected error rate across all locations. The global precision, $M_0$, estimates the variation in the error rate  across locations. The local precision, $M_j$, estimates the variation in the error rate across replicates at location $j$.

@@

Figure~\ref{fig:graphical_model} shows a graphical representation of the RVD2 statistical model. In this graphical model framework a shaded node represents an observed random variable, an unshaded node represents an unobserved or latent random variable and a directed edge represents a functional dependency between the two connected nodes~\citep{Jordan2004}. A rounded box or ``plate" represents replication of the nodes within the plate. The graphical model framework connects graph theory and probability theory in a way that facilitates algorithmic methods for statistical inference.

@@

\begin{equation}\label{eqn:jointpdf}
p \left( r_{ji}, \theta_{ji}, \mu_j | n_{ji}; \mu_0, M_0, M_j \right) = p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right),
\end{equation}

where
\begin{align}
p\left( \mu_j; \mu_0, M_0 \right)  &= \frac{ \Gamma(M_0) } { \Gamma(\mu_0 M_0) \Gamma(M_0 (1-\mu_0)) } \mu_j^{M_0\mu_0 -1} (1 - \mu_j)^{M_0 ( 1 - \mu_0) - 1}, \nonumber \\
p\left( \theta_{ji} | \mu_j; M_j \right) &= \frac{ \Gamma(M_j) } { \Gamma(\mu_j M_j) \Gamma(M_j (1-\mu_j)) } \theta_{ji}^{M_j\mu_j -1} (1 - \theta_{ji})^{M_j ( 1 - \mu_j) - 1}, \nonumber \\
p \left( r_{ji} | \theta_{ji}, n_{ji} \right) &= \frac{ \Gamma(n_{ji}+1) } { \Gamma(r_{ji}+1) \Gamma( n_{ji} - r_{ji} + 1 ) } \theta_{ji}^{r_{ji}} (1 - \theta_{ji})^{n_{ji} - r_{ji}}. \nonumber
\end{align}

Integrating over the latent variables $\theta_{ji}$ and $\mu_j$ yields the marginal distribution of the data, 
\begin{equation}
p \left( r_{ji} | n_{ji} ; \mu_0, M_0, M_j \right) = \int_{\mu_j} \int_{\theta_{ji}}  p \left( r_{ji} | \theta_{ji}, n_{ji} \right) p\left( \theta_{ji} | \mu_j; M_j \right) p\left( \mu_j; \mu_0, M_0 \right) d\theta_{ji} d\mu_j.
\end{equation}

Finally, the log-likelihood of the data set is

@@

The Beta distribution over $\mu_j$ is conjugate to the Binomial distribution over $\theta_{ji}$, so we can write the posterior distribution as a Beta distribution. However, there is not a closed form for the product of a Beta distribution with another Beta distribution, so exact inference is intractable.

@@

We present the initial parameter estimates here and provide the derivations in Appendix~\ref{sec:appendix_mom}. The MoM estimate for replicate-level parameters are 
$\tilde{\theta}_{ji} = \frac{r_{ji}} {n_{ji}}$. 
The estimates for the position-level parameters are 
$\tilde{\mu}_j = \frac{1}{N} \sum_{i=1}^N \theta_{ji}$ 
and 
$\tilde{M_j} = \frac{ \tilde{\mu}_j (1 - \tilde{\mu}_j ) } { \frac{1}{N} \sum_{i=1}^N \theta_{ji}^2 } -1$. 
The estimates for the genome-level parameters are 
$\tilde{\mu}_0 = \frac{1}{J} \sum_{j=1}^J \mu_j$ 
and 
$\tilde{M}_0 = \frac{ \tilde{\mu}_0 (1 - \tilde{\mu}_0 ) } {\frac{1}{J} \sum_{j=1}^J \mu_j^2 } -1$. 
While MoM is known to have the potential to return estimates outside the true parameter bounds, we have not encountered such behavior in this application.

@@

%%%%%%%%%%%%%%%%%%
% Sampling mu
%%%%%%%%%%%%%%%%%%
\subsection{Sampling from $p \left( \mu_j |\theta_{ji},M_j,\mu_0,M_0\right)$}
The posterior distribution over $\mu_j$ given its Markov blanket is 
\begin{equation}
	p( \mu_j | \theta_{ji}, M_j, \mu_0, M_0 ) \propto p(\mu_j | \mu_0, M_0) p(\theta_{ji} | \mu_j, M_j).
\end{equation}

@@

A candidate sample is generated from the symmetric proposal distribution $Q(\mu_j^* | \mu_j^{(p)}) \thicksim \mathcal{N}(\mu_j^{(p)}, \sigma_j^2)$, where $\mu_j^{(p)}$ is the $p$th from ythe posterior distribution. The acceptance probability is then
\begin{equation}
	a = \frac{ p(\mu_j^* | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^*, M_j) } {p(\mu_j^{(p)} | \mu_0, M_0) p(\theta^{(p+1)}_{ji} | \mu_j^{(p)}, M_j)}
\end{equation}

We fixed the proposal distribution variance for all the Metropolis-Hastings steps within a Gibbs iteration to $\sigma_j^2 = 0.1 \cdot \mu_j^{(p)}$ if $\mu_j^{(p)} >0$ and $0.1$ otherwise. Though it is not theoretically necessary, we have found that the algorithm performance improves when we take the median of five or more M-H samples as a single Gibbs step for each position. 
%TODO: make sure \sigma_j is correct here.

We resample from the proposal if the sample is outside of the support of the posterior distribution. We typically discard 20\% of the sample for burn-in and thin the chain by a factor of 2 to reduce autocorrelation among samples. Since, each position $j$ is exchangeable given the global hyperparameters $\mu_0$ and $M_0$ this sampling step can be distributed across up to $J$ processors.

%%%%%%%%%%%%%%%%%%
% Posterior Density Test
%%%%%%%%%%%%%%%%%%
\subsection{Posterior Density Test}\label{sec:hypothesis_test}
Metropolis-within-Gibbs provides samples from the posterior distribution of $\mu_j$ given the case or control data. For notational simplicity, we define the random variables associated with these two distributions $\tilde{\mu}_j^{\text{case}}$ and $\tilde{\mu}_j^{\text{control}}$.

@@

The threshold, $\tau$, may be set to zero or optimized for a given median depth and desired MAF detection limit. The optimal $\tau$ minimizes the sum of the false negative rate and false positive rate or, equivalently,  the $L_1$ distance to perfect classification in the ROC curve plot,
\begin{equation}
	\tau^* = \arg\min_\tau \left\{ (1-\text{TPR}(\tau)) + \text{FPR}(\tau) \right\}.
\end{equation}

While we are able to compute the optimal $\tau$ threshold for a test data set, in general we would not have access to $\tau^*$. With sufficient training data, one would be able to develop a lookup table or calibration curve to set $\tau$ based on read depth and MAF level of interest. Absent this information we set $\tau = 0$.

@@

An abundance of non-reference bases at a position called by the posterior density test may be due to a true mutation or due to a random sequencing error; we would like to differentiate these two scenarios. We assume non-reference read counts caused by a non-biological mechanism results in a uniform distribution over three non-reference bases. In contrast, the distribution of counts among three non-reference bases caused by biological mutation would not be uniform.

@@

where $r_{ji}^{(k)}$ is the observed frequency for non-reference base $k$ at position $j$ in replicate $i$ and $E_{ji}^{(k)}$ is the corresponding expected frequency under the null hypothesis. \citet{Cressie1984} recommended $\lambda = 2/3$ when no knowledge of the alternative distribution is available and we choose that value.

We control for multiple hypothesis testing in two ways. We use Fisher's combined probability test \citep{} to combine the p-values for $N$ replicates into a single p-value at position $j$,

\begin{equation}\label{eqn:fisher_combined}
	X_j^2 = -2 \sum_{i=1}^N \ln(p_{ji}).
\end{equation}

Equation~\eqref{eqn:fisher_combined} gives a test statistic that follows a $\chi^2$ distribution with $2N$ degrees of freedom when the null hypothesis is true. Finally, we use the Bejamini-Hochberg method to control the family-wise error rate (FWER) over positions that have been called by the Bayesian hypothesis test~\eqref{eqn:bayes_test} \citep{}.
%TODO: add two citations here

@@

We used two independent data sets to evaluate the performance of RVD2 and compare it with other variant calling algorithms. Synthetic DNA sequence data provides true positive and true negative positions as well as define minor allele fractions. HCC1187 data is used to test the performance on a sequenced cancer genome with less than 100\% tumor purity.

@@

Two 400bp DNA sequences that are identical except at 14 loci with variant bases were synthesized and clonally isolated and labeled case and control. Sample of the case and control DNA were mixed at defined fractions to yield defined minor allele frequencies (MAFs) of 0.1\%, 0.3\%, 1\%, 10\%, and 100\%. More details of the experimental protocol are available from the original publication~\citep{Flaherty:2011ja}. We aligned the reads to the reference sequence using BWA v0.7.4 with the -C50 option to filter for high mapping quality reads.

@@

The HCC1187 cell line is derived from  epithelial cells from primary breast tissue from a 41 y/o adult with TNM stage IIA primary ductal carcinoma. A matched normal cells were derived from lymphoblastoid cells from whole blood~\citep{}. 
% TODO: cite the original publication here.
% TODO: check matched cells line origin
% double checked by Yuting, didn't find the matched cells line origin
% TODO: mention tumor purity here

@@

We generated receiver-operating characteristic curves (ROCs) for a range of median read depth and a range of minor allele frequencies (MAFs). For these ROC curves, we used the Bayesian test without the $\chi^2$ test. Adding the $\chi^2$ improves specificity at the expense of sensitivity. Figure~\ref{fig:ROC} shows ROC curves generated by varying the threshold $\tau$ with a fixed $\alpha=0.05$. Figure~\ref{fig:ROC}A shows ROC curves for a true 0.1\% MAF for a range of median coverage depths. At the lowest depth the sensitivity and specificity is no better than random. However, we would not expect to be able to call a 1 in 1000 variant base with a coverage of only 43. The performance improves monotonically with read depth. Figures~\ref{fig:ROC}B-C show a similar relationship between coverage depth and accuracy for higher MAFs.

@@

%Figure~\ref{fig:MAF} shows the posterior mean and 95\% credible intervals for $\mu_j$ for called variant positions with $\bar{n} = 5584$ and MAF = 1.0\%. All of the called positions show a clear difference between the case and control error rates. The posterior mean estimates are all shrunken towards the global error rate parameter $\mu_0 = 0.0023$ due to the hierarchical structure of the model.
%
%\begin{figure}[h]
%\begin{center}
%\includegraphics[width=120mm]{pdf_figs/Synthetic_Mubarplot_MAF1_0_Depth5584.pdf}
%\caption{Estimated minor allele fraction for called variants in 1.0\% dilution.}
%\label{fig:MAF}
%\end{center}
%\end{figure}
%% from folder '2013-11-15_synthetic_mu_barplot_optimalT', dilution rate is 0.1, downsampling rate is 0.01.

@@

We compare the empirical performance of RVD2 to other variant calling algorithms using the synthetic DNA data sets using the false discovery rate as well as sensitivity/specificity. In a research applications, the false discovery rate is a more relevant performance metrics because the aim is generally to identify interesting variants. The sensitivity/specificity metric is more relevant in clinical applications where one is more interested in correctly calling all of the positive variants and none of the negatives. GATK, Varscan2, strelka and muTect are only able to make use of one case and cone control sample, so we provide results of RVD2 with the same data ($N=1$) for a fair comparison.

@@

The sensitivity for RVD2 with $\tau=0$ is low for low read depths and MAF levels and $N=1$ case and control sample. The sensitivity increases considerably with read depth at a slight expense to specificity. With $\tau^*$ the performance is much better with high sensitivity and specificity across a wide range of read depths and MAFs. However, in practice one may not know the optimal $\tau^*$ a-priori. With $N=6$ replicates, the sensitivity increases considerably for low MAF variants with a slight degradation in specificity due to false positives. When the median read depth is at least $10\times$ the MAF, RVD2 has higher specificity than all of the other algorithms tested and has a lower sensitivity in only three cases. The Matthews correlation coefficient (MCC) indicates that RVD2($\tau=0, N=1$) is more accurate than the other algorithms when the median read depth is at least $10\times$ the MAF (see Appendix~\ref{sec:app_comparison}).

@@

VarScan2-somatic is able to make calls for all but the lowest MAF and coverage level. However, the FDR is high due to many false positives. Interestingly, at a MAF of 100\% the FDR is zero for lowest read depth and over 0.9 for the highest read depth. Strelka has a better FDR than the samtools, GATK or Varscan2-somatic algorithms for almost all read depths at the 10\% and 100\% MAF. However, it does not call any variants at or below 1\% MAF.  MuTect has the best FDR performance of the other algorithms we tested over a wide range of MAF and depths. But the FDR level is relatively high at around 0.7 for 0.1\% -- 1\% MAF and 0.3 for 10\% -- 100\% MAF.

@@

RVD2 identified five variants in the 44kbp PAXIP1 gene from chr7:XXX to chr7:XXX. Of the five called variants, three were homozygous germline variants and two were heterozygous germline variants. The two heterozygous variants had a loss-of-heterozygosity in the tumor sample. 
%TODO: add start and stop location here

Figure~\ref{fig:brca_MAF} shows the estimated minor allele frequencies for the normal and tumor samples at the called locations. Positions chr7:154743899C$>$T, chr7:154753635T$>$C and chr7:154780960C$>$T are called germline homozygous mutations. Positions chr7:154754371 and chr7:154758813 are called heterozygous in the normal sample. In the tumor sample we identified five mutations chr7:154743899C$>$T, chr7:154753635T$>$C, chr7:154754371T$>$C, chr7:154758813G$>$A, and chr7:154780960C$>$T. Positions chr7:154754371T$>$C and chr7:154758813G$>$A are called loss-of-heterozygosity events. Some of these mutations are also found to be common population SNPs according to dbSNPv138. The corresponding identities are shown in the figure.
%TODO: verify these rs positions
%According to dbSNPv138 chr7:154743899 is rs1239326, chr7:154753635 is rs1239324, chr7:154754371 is rs71534174, chr7:154758813 is rs35505514, and chr7:154780960 is rs4398858.

@@

The original research describing this sample used Strelka to identify mutations. They identified chr7:154760439 as variant, but did not call the other four variants. In particular Strelka missed the two LOH events.

@@

We describe here a novel algorithm for model estimation and hypothesis testing for identifying single-nucleotide variants in heterogeneous samples using next-generation sequencing data. Our algorithm has a higher sensitivity and specificity than many other approaches for a range of read depths and minor allele frequencies.

Our inference algorithm uses Gibbs sampling to estimate the hierarchical empirical Bayes model. This sampling procedure provides a guarantee to identify the global optimal parameter settings asymptotically. However, it may require many samples to achieve that guarantee causing the algorithm to be slower than other deterministic approaches. We opted for this balance of speed and accuracy because computational time is often not limiting and the loss of a false positive or false negative greatly outweighs the cost of more computation.

@@

%%%%%%%%%%%%%%%%%%%
% Appendix B: Comparison Statistics
%%%%%%%%%%%%%%%%%%%
\section{Algorithm Comparison Statistics}\label{sec:app_comparison}
%TODO: discuss this result

@@

%%%%%%%%%%%%%%%%%
% Appendix D: Parameter Settings for Variant Calling Algorithms
%%%%%%%%%%%%%%%%%
\section{Parameter settings for other variant calling algorithms}
%TODO: review grammar and completeness here
